Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks

### 2025-05-06 inst/paper/main.tex
add info on serial interval from skew-normal

To assess the efficacy of isolation and contact tracing to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025}. The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (without depletion of susceptibility). The mean of the negative binomial offspring distribution is the basic reproduction number, and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}. \\

The NPI implemented in the transmission model is an isolation of symptomatic cases. The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation). All infection times that are sampled to occur after the isolation of an infector are not infected. Therefore, the isolation of a case can reduce the number of secondary cases if isolation happens prior to any of the secondary infection times. Another way to iterpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).  \\

@@

We define the proportion of pre-symptomatic transmission follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skewed-normal distribution to sample the infection times using the incubation period and a skew parameter ($\theta$). The skew parameter controls the proportion of pre-symptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:prop-presym-trans}). We parameterised the skew-normal distribution with a skew of 0.7, 1.95 and 30, these correspond to a pre-symptomatic transmission percentage of 30\%, 15\% and $<1\%$, respectively. Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective contact tracing is hypothesied to be. \\

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/prop_presymptomatic_transmission.png}
\caption{Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skewed-normal distribution.}
\label{fig:prop-presym-trans}
\end{figure}

>>>>

To assess the efficacy of isolation and contact tracing to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025}. The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (without depletion of susceptibility). The mean of the negative binomial offspring distribution is the basic reproduction number ($R$), and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}. \\

There are three categories for individuals in the simulation, with each having their own parameterisation for the negative binomial offspring distribution: 1) \textit{community} (i.e. general population), 2) \textit{isolated}, and 3) \textit{asymptomatic}. Each group have a distinct $R$ and dispersion ($k$) parameters due to the assumed differences in transmission dynamics between groups. \\

The time between an infector being infected and then infecting an infectee (i.e. serial interval) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}. This controls the proportion of the infections that occur prior to, or after, the symptom onset of an infector. The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed. The scale parameter ($\omega$) is fixed at two, this parameter (with the shape parameter) defines the variance of the distribution. We select this value as to provides a realistic amount of variance when transforming onset time to exposure time (variance range 1.5 - 3.2). The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect. Overall, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we compute the time an infectee is exposed/infected. \\

The NPI implemented in the transmission model is an isolation of symptomatic cases. The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation). All infection times that are sampled to occur after the isolation of an infector are not infected. Therefore, the isolation of a case can reduce the number of secondary cases if isolation happens prior to any of the secondary infection times. Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).  \\

@@

We define the proportion of pre-symptomatic transmission ($\theta$) follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skew-normal distribution to sample the infection times using the incubation period and a shape parameter ($\alpha$). The shape parameter controls the proportion of pre-symptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:prop-presym-trans}). We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: <1\%, 15\%, 30\%, these correspond to $\alpha$ values of 30, 1.95 and 0.7, respectively (Figure \ref{fig:prop-presym-trans}). Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective contact tracing is hypothesied to be. \\

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/prop_presymptomatic_transmission.png}
\caption{Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skew-normal distribution.}
\label{fig:prop-presym-trans}
\end{figure}
